APOGEE CATALOG OF 10,000+ M DWARFS MEASURED WITH DATA-DRIVEN MODELS

### 2018-07-30 paper/draft1/draft1.tex
update draft

\documentclass[modern]{aastex62}

@@

\title{APOGEE CATALOG OF 10,000+ M DWARFS MEASURED BY DATA-DRIVEN MODELS}

@@

% \begin{enumerate}
Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}.

@@

Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.

 (1) model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\citealt{Rajpurohit:2014}, \citealt{Rajpurohit:2018}) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\citealt{Allard:2013})--models don't accurately reproduce spectra plus unquantified systematic errors;

 (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (\citealt{Casagrande:2008} estimated effective temperatures, bolometric luminosities, and metallicities using optical and infrared photometries) -- however we want to use \emph{all} of the data!

 (3) using empirical classifications/templates/standards - SPT, luminosity class - this is a relative system but does not provide an absolute standard for measuring physical parameters

@@

 (1) Can be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects;

 (2) Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all

 (3) better than template matching in that it is using less sparsely sampled known mdwarfs

@@

The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III (SDSS-III; \citealt{Eisenstein:2011}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}). Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).

APOGEE is primarily targeted for bright stellar populations, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution. However numerous cool, main sequence sources have also been observed either accidentally, or as targets proposed by the M dwarf ancilliary survey \citep{Desphande:2013} (specific estimate based on Gaia?).

@@

Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector \emph{l} and and coefficient vector \emph{$\theta$}. So for each star \emph{n}, at pixel wavelength \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
\begin{equation}
	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + {\rm noise}
\end{equation}
We use the noise model is ${\rm noise}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.

@@

Another mode of analysis we can look at with The Cannon is how the derivative of the model changes with respect to given training parameter, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters. As discussed further in sections \ref{subsec:mann_results} and \ref{subsec:west_results}, we show the derivative plots of each of our two models and compute the error of the derivative at each pixel using a jackknife statistic:
	\[\sigma_{\theta,m}^2 = \frac{N-1}{N} \sum^N_{n=1} (\theta_{/n}^m - \theta^m)^2 \]
where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample, indexed by n, $\theta$ is the coeffient vector trainined on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.

\subsubsection{Temperature/Metallicity Model \label{subsec:mann_results}}
For the physical parameter model we trained The Cannon on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 68K/0.07dex respectively, similar to the original training sample uncertainties of 60K/0.08dex. We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$Gyr) and likely slowly rotating sources. While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources does not significantly

@@

\subsubsection{Spectral Type Model \label{subsec:west_results}}
We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type. We note however, that the training sample is distributed heavily towards earlier type sources, with only one M8 and one M9 source.

\subsection{Test Sample}
Out of the total APOGEE DR14 catalog of 277,371 sources, we selected 254,478 sources which were in the cross match of Gaia DR2 (\citealt{Brown:2018}) and applied Gaia color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources. From there we applied the following selection procedure to identify a "safe" sample of M dwarfs:

@@

\begin{enumerate}
\item[-] Brief summary of what we did and found

@@

\item[-] Some discussion of whether the derivatives are sensible or surprising. \color{red}{Add derivative plots.}\color{black}

\item[-] The Cannon as a test of evaluating training parameter accuracy? 
\item[-] Discussion of validation tests and ground truth
\end{enumerate}

%%==========================================================================================
\acknowledgements
We would like to acknowledge Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Hans Walter Rix (MPIA), Anna-Christina Eilers (MPIA) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality. \color{gcolor}{HOGG: SDSS acknowledgments... Grants and stuff...}\color{black}

@@

%---------------------------
\begin{figure}[ht]
\plotone{figures/Spectral_Sequence_1.pdf}
\plotone{figures/Spectral_Sequence_2.pdf}
\plotone{figures/Spectral_Sequence_3.pdf}
\caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
\end{figure}

@@

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/derivative_jackknife_teff.png}
\includegraphics[width=16cm]{figures/derivative_jackknife_feh.png}
\end{center}
\caption{Derivative plots for physical parameter model. \textit{Top:} Derivative of flux with repsect to temperature taken at the median training temperature, T$_{\rm eff}=3463K$. \textit{Bottom:} Derivative of flux with repsect to metallicity taken at the median training metallicity, [Fe/H]$=-0.03$ dex.} \label{fig:mann_derivative}
\end{figure}

%---------------------------
\begin{figure}[ht]
\plotone{figures/cannon_chi_dist.png}
\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE-Gaia cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$. We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as "safe". \label{fig:chi_dist}}
\end{figure}

\begin{figure}[ht]
\plotone{figures/cmd_selection.png}
\caption{Color-magnitude cuts.}
\end{figure}

\begin{figure}[ht]
\plotone{figures/safe_selection.png}
\caption{Safe sample selection.}
\end{figure}

%---------------------------
\begin{figure}[ht]
\plotone{figures/aspcap_cannon_validation.png}
\caption{Validation test}
\plotone{figures/aspcap_cannon_label_hist.png}
\caption{Distribution of temperatures and metallicities from The Cannon compared to the ASPCAP pipeline.}
\end{figure}

>>>>

\documentclass[preprint]{aastex62}

@@

\title{APOGEE CATALOG OF 10,000+ M DWARFS MEASURED WITH DATA-DRIVEN MODELS}

@@

% \begin{enumerate}
Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}.

@@

Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled. Several approaches to to determining physical parameters and the limitations:

\begin{enumerate}
\item model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\citealt{Rajpurohit:2014}, \citealt{Rajpurohit:2018}) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\citealt{Allard:2013})--models don't accurately reproduce spectra plus unquantified systematic errors;

\item model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (\citealt{Casagrande:2008} estimated effective temperatures, bolometric luminosities, and metallicities using optical and infrared photometries) -- however we want to leverage the full wavelength range and resolution of our data.

\item using empirical classifications/templates/standards - SPT, luminosity class - this is a relative system but does not provide an absolute standard for measuring physical parameters
\end{enumerate}

@@

\begin{enumerate}
\item Can be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects;

\item Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all

\item better than template matching in that it is using less sparsely sampled known mdwarfs
\end{enumerate}

@@

The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (SDSS-III/IV \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}). Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).

APOGEE is primarily targeted for bright stellar populations, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution. However numerous cool, main sequence sources have also been observed either as targets proposed by the M dwarf ancilliary survey (1,350 sources selected by proper motions from the  \citealt{Desphande:2013}), or randomly by APOGEE, which we quantify in this study to be over 10,000 in total.

@@

Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector \emph{l} and and coefficient vector \emph{$\theta$}. So for each star \emph{n}, at pixel wavelength \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
\begin{equation}
	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + {\rm noise}
\end{equation}
We use the noise model is ${\rm noise}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.

@@

Another mode of analysis we can look at with The Cannon is how the derivative of the model changes with respect to given training parameter, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters. As discussed further in sections \ref{subsec:mann_results} and \ref{subsec:west_results}, we show the derivative plots of each of our two models and compute the error of the derivative at each pixel using a jackknife statistic:
\begin{equation}
	\sigma_{\theta,m}^2 = \frac{N-1}{N} \sum^N_{n=1} (\theta_{/n}^m - \theta^m)^2 
\end{equation}
where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample, indexed by \emph{n}, $\theta$ is the coeffient vector trainined on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.

\subsubsection{Temperature/Metallicity Model \label{subsec:mann_results}}
For the physical parameter model we trained The Cannon on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 68K/0.07dex respectively, similar to the original training sample uncertainties of 60K/0.08dex. We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr) and likely slowly rotating sources. While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.

@@

\subsubsection{Spectral Type Model \label{subsec:west_results}}
We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type. We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source. As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates The Cannon does a poor job at extrapolating to labels outside of the training sample space.

\color{gcolor}{Further assessment of the reliability of our model for late types?...}\color{black}

\subsection{Test Sample \label{subsec:test_selection}} 
Out of the total APOGEE DR14 catalog of 277,371 sources, we selected 254,478 sources which were in the cross match of Gaia DR2 (\citealt{Brown:2018}) and applied Gaia color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources. From there we applied the following selection procedure to identify a sample of single, main-sequence M stars. Although our catalog reports the temperatures, metallicities, and spectral types for XXX sources out of 14,000 with $\chi^2<200,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):

\begin{enumerate}
\item \textbf{Known contaminant cut:} We remove known pre-main sequence sources from \citealt{Cottaar:2014} (280 sources), and known binaries from \citealt{ElBadry:2018} and \citealt{Skinner:2018} (58 sources).

\item \textbf{K dwarf cut:} In 2MASS photometry we select $M_K > 4.3$ to remove likely K dwarfs, which appear reddened in Gaia $BP$, $RP$ photometry.

\item \textbf{Pre-main sequence, multiple system and subdwarf (?) cut:} in Gaia and 2MASS we apply $.7<J-K<1.1$ and points in the selction region within the points $\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ of ($BP-RP$, $M_G$) to cut remaining likely pre-main sequence, multiple sources and subdwarfs.

\item \textbf{Quality of fit cut:} We apply a $\chi^2$ cut of less than 100,000.

\item \textbf{Training bound selection:} Because extrapolation with The Cannon is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<[Fe/H]<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
\end{enumerate}

The final selection for our ``safe'' is displayed in Figure \ref{fig:safe_selection}. We also note that applying our model requires very little computational demand: the time to run a model on all 14,827 sources was two minutes on a personal laptop.

In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of ASPCAP temperature values.

@@

\begin{enumerate}
\item[-] Brief summary of what we did and found{}

@@

\item[-] Some discussion of whether the derivatives are sensible or surprising.

\item[-] The Cannon as a test of evaluating training parameter accuracy? \color{gcolor}{HOGG: Should this be precision, not accuracy? For example adding a constant factor (say adding +1000 to every metallicity training value) does not change the scatter of the test labels}\color{black}

\item[-] Discussion of validation tests and ground truth
\end{enumerate}

%%==========================================================================================
\acknowledgements
We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality. \color{gcolor}{HOGG: SDSS acknowledgments... Grants and stuff...}\color{black}

@@

@@

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=12cm]{figures/Spectral_Sequence_1.png}
\includegraphics[width=12cm]{figures/Spectral_Sequence_2.png}
\includegraphics[width=12cm]{figures/Spectral_Sequence_3.png}
\end{center}
\caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
\end{figure}

@@

% \begin{figure}[ht]
% \begin{center}
% \includegraphics[width=16cm]{figures/derivative_jackknife_teff.png}
% \includegraphics[width=16cm]{figures/derivative_jackknife_feh.png}
% \end{center}
% \caption{Derivative plots for physical parameter model. \textit{Top:} Derivative of flux with repsect to temperature taken at the median training temperature, T$_{\rm eff}=3463K$. \textit{Bottom:} Derivative of flux with repsect to metallicity taken at the median training metallicity, [Fe/H]$=-0.03$ dex.} \label{fig:mann_derivative}
% \end{figure}

\begin{figure}[ht]
\plotone{figures/cannon_chi_dist.png}
\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE-Gaia cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$. We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as "safe". \label{fig:chi_dist}}
\end{figure}

\begin{figure}[ht]
\plotone{figures/cmd_selection.png}
\caption{Gaia and 2MASS color-magnitude cuts for 12,037 sources with $\chi^2<100,000$. Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by Gaia. \label{fig:cmd_selection}}
\end{figure}

\begin{figure}[ht]
\plotone{figures/safe_selection.png}
\caption{Final sample selection with 10,084 M dwarfs based on criteria described in Section \ref{subsec:test_selection}. \label{fig:safe_selection}}
\end{figure}

\begin{figure}[ht]
\plotone{figures/aspcap_cannon_label_hist.png}
\caption{Distribution of temperatures and metallicities for 10,084 sources from The Cannon compared to the ASPCAP pipeline. \label{fig:aspcap_cannon_label_hist}}
\end{figure}

%---------------------------
\begin{figure}[ht]
\plotone{figures/aspcap_cannon_validation.png}
\caption{Validation test \label{fig:aspcap_cannon_validation}}
\end{figure}
